---
output:
  md_document:
    variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(gnomonicM)
```

# gnomonicM
[![packageversion](https://img.shields.io/badge/Package%20version-1.0.1-orange.svg?style=flat-square)](commits/master)  [![metacran downloads](http://cranlogs.r-pkg.org/badges/gnomonicM)](http://cran.rstudio.com/web/packages/gnomonicM/index.html) [![minimal R version](https://img.shields.io/badge/R%3E%3D-3.6.0-6666ff.svg)](https://cran.r-project.org/)

**Estimate Natural Mortality (M) throughout the life history of species**
This package allows to estimate Natural mortality for different life stages for organism, usually fish and invertebrates, based on the gnomonic interval approach (Caddy, 1991, 1996; Martínez-Aguilar et al., 2005). We have included improvements modifying some equations and the estimation procedure.


## Installation

Get the development version from github:

```R
# install.packages("devtools")
devtools::install_github("ejosymart/gnomonicM")
```

Or install the CRAN version:

```R
install.packages("gnomonicM")
```


After, that call the package:
```R
library("gnomonicM")
```


## Examples

This is a basic example which shows you how to estimate natural mortality based on gnomonic approach using the data on Caddy (1996):

## Deterministic method

```{r}
model <- gnomonic(nInterval   = 7, 
                  eggDuration = 2, 
                  longevity   = 365, 
                  fecundity   = 200000, 
                  a_init      = 2)

print(model)
```


## Stochastic method

```{r}
modelUniform <- gnomonicStochastic(nInterval     = 7, 
                                   eggDuration   = 2,
                                   longevity     = 365,
                                   distr         = "uniform", 
                                   min_fecundity = 100000, 
                                   max_fecundity = 300000, 
                                   niter         = 1000, 
                                   a_init        = 2)
```


For more details, please read the vignettes of this package.


### References

Caddy JF (1991). Death rates and time intervals: is there an alternative to the constant natural mortality axiom? Reviews in Fish Biology and Fisheries 1:109–138. DOI: 10.1007/BF00157581.

Caddy JF (1996). Modelling natural mortality with age in short-lived invertebrate populations: definition of a strategy of gnomonic time division. Aquatic Living Resource 9:197–207. DOI: 10.1051/alr:1996023.

Martínez-Aguilar S, Arreguín-Sánchez F, Morales-Bojórquez E (2005). Natural mortality and life history stage duration of Pacific sardine (*Sardinops caeruleus*) based on gnomonic time divisions. Fisheries Research 71:103–114. DOI: 10.1016/j.fishres.2004.04.008.
